---
title: "Pacific centered map and Voronoi tessellation"
author: "by: [Valentin Stefan](https://github.com/valentinitnelav) "
date: "last update: `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    fig_caption: true
    number_sections: true
    theme: "journal"
    highlight: "tango"
---

# Summary

The main steps for this map consist in getting country polygons data, cropping and shifting the geographical coordinates for a Pacific view, labeling some entities (here some small landmasses in Pacific) and adding boundaries between them using Voronoi tessellation method.

The Rmarkdown file used to generate this html report can be found [here](https://github.com/valentinitnelav/valentinitnelav.github.io/tree/master/assets/2020-05-15-pacific-map-voronoi/pacific-map-voronoi.Rmd).

# The R environment {.tabset .tabset-fade}

## Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(rgeos)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(ggplot2)
library(ggrepel)
```

## R Session

```{r r-session-info}
sessionInfo()
```

## Package versions

If you want to install a snapshot of the packages as they existed on CRAN at the date of the creation of this document, then can use the [checkpoint][cp] package. The code below should guarantee the same versions of the packages as they existed on CRAN at the specified point in time.

[cp]: https://cran.r-project.org/web/packages/checkpoint/index.html

```{r checkpoint, eval=FALSE}
library(checkpoint) # checkpoint, version 0.4.9
checkpoint(snapshotDate = "2020-05-16")
# Check that library path is set to ~/.checkpoint/2020-05-16/ ...
.libPaths()
grepl(pattern = "\\.checkpoint/2020-05-16/", x = .libPaths()[1]) # should be TRUE
```

The `checkpoint` functionality consists on scanning for package names in the scripts and text files of your project folder and its subfolder. It scans all R code (.R, .Rmd, and .Rpres files) for `library()` and `require()` statements. Then creates a local library into which it installs a copy of the packages required in the project as they existed on CRAN at the specified snapshot date. See details with `?checkpoint` after you load `library(checkpoint)`, or [here](https://cran.r-project.org/web/packages/checkpoint/vignettes/checkpoint.html)

**Warning** - Installing older versions of the packages in the checkpoint local library, may take up some hundreds of MG in space.

# Data preparation

## Set ggplot2 theme

```{r, ggplot2-theme}
theme_set(
  theme_bw() +
    theme(panel.background = element_rect(fill = "azure"),
          panel.grid.major = element_blank(),
          axis.title = element_blank(),
          axis.text = element_text(size = 8))
)
```

## Natural Earth data

Get world spatial polygons and focus on the Pacific area.
```{r, world-data}
world <- rnaturalearth::ne_countries(scale = 'medium', returnclass = "sp")
box_cut <- bbox2SP(n = 90, s = -90, w = -70, e = 120, proj4string = world@proj4string)
world_crop <- gDifference(world, box_cut)
```

The main steps consist in cutting out the read area and binding the two green ones with shifting the geographical coordinates for a Pacific view as illustrated below:
```{r crop-explanation, echo=FALSE}
box_left <- bbox2SP(n = 90, s = -90, w = -180, e = -70, proj4string = world@proj4string)
box_right <- bbox2SP(n = 90, s = -90, w = 120, e = 180, proj4string = world@proj4string)

plot(world)
plot(box_cut, col = rgb(red = 1, green = 0, blue = 0, alpha = 0.5), add = TRUE)
plot(box_left, col = rgb(red = 0, green = 1, blue = 0, alpha = 0.5), add = TRUE)
plot(box_right, col = rgb(red = 0, green = 1, blue = 0, alpha = 0.5), add = TRUE)

plot(world_crop)
plot(box_left, col = rgb(red = 0, green = 1, blue = 0, alpha = 0.5), add = TRUE)
plot(box_right, col = rgb(red = 0, green = 1, blue = 0, alpha = 0.5), add = TRUE)

pacific <- world_crop %>% 
  st_as_sf() %>% 
  st_shift_longitude()
plot(pacific)
```

Next we will further crop the results from above to a focus area on the Pacific depicted in green below:
```{r, crop-2nd, echo=FALSE}
plot(pacific)
bbox2SP(n = 30, s = -50,
        w = st_bbox(pacific)[["xmin"]],
        e = st_bbox(pacific)[["xmax"]],
        proj4string = world@proj4string) %>% 
  plot(col = rgb(red = 0, green = 1, blue = 0, alpha = 0.5), add = TRUE)
```

So, we take the world polygons obtained by cutting out the central box, shift the geographical coordinates for a Pacific view and crop to focus on a more narrow area:
```{r, pacific_crop}
pacific_crop <- world_crop %>% 
  st_as_sf() %>% 
  st_shift_longitude() %>% 
  st_crop(c(xmin = st_bbox(.)[["xmin"]],
            xmax = st_bbox(.)[["xmax"]],
            ymin = -50,
            ymax = 30))

ggplot() +
  geom_sf(data = pacific_crop)
```

For the labeling exercise we can use the `rnaturalearthdata::tiny_countries50` dataset. This set of polygons also need the be shifted and cropped.
```{r, tiny-countries}
tiny_countries <- rnaturalearthdata::tiny_countries50 %>% 
  st_as_sf() %>% 
  st_shift_longitude() %>% 
  st_crop(c(xmin = 120, xmax = 250, ymin = -50, ymax = 30)) %>% 
  # Also adds the coordinates to be used for labeling with geom_text_repel
  bind_cols(st_coordinates(.) %>% as.data.frame())

ggplot() +
  geom_sf(data = pacific_crop) +
  geom_sf(data = tiny_countries, size = 2, color = "red")
```

## Voronoi tessellation

This will create boundaries between the land masses we decided to label. Cropping is needed here, because otherwise we get very long segments at the edges of the Voronoi grid.

```{r, voronoi}
voronoi_lines <- tiny_countries %>%
  st_as_sf() %>% 
  st_union() %>% # this seems to be needed, otherwise st_voronoi will fail
  st_voronoi(bOnlyEdges = TRUE) %>% 
  st_crop(st_bbox(pacific_crop))
```
Note - ignore the warnings. Things do not have to be perfect here, we use the Voronoi grid for visual effects only and you will see that it gets the job done.

# The map

Put all together in a map

```{r, make-map}
pacific_map_voronoi <- ggplot() +
  geom_sf(data = voronoi_lines, color = gray(level = 0.6), size = 0.5) +
  geom_sf(data = pacific_crop, size = 0.25) +
  geom_sf(data = tiny_countries, color = "black", size = 0.5) +
  # Here, coord_sf ensures that all layers are displayed within the same limits,
  # the extremes being dictated by the minimum and maximum of pacific_crop. Also
  # the expanding factor is turned off so that the content of the map can touch
  # the axes:
  coord_sf(xlim = c(120, 290), expand = FALSE) +
  geom_text_repel(data = tiny_countries,
                  aes(x = X, y = Y, label = name),
                  size = 2.5,
                  fontface = "bold",
                  segment.size = 0.25,
                  box.padding = 0.4,
                  min.segment.length = 0,
                  seed = 2020-5-16)
pacific_map_voronoi
```

If you need to further save the map as a png file, you can, of course, use `ggsave()`:
```{r}
ggsave(filename = "pacific_map_voronoi.png",
       plot = pacific_map_voronoi,
       width = 15,
       height = 10,
       units = "cm",
       dpi = 300)
```

